---
title: "Human Activity Research: Dumbbell Exercise Course Project"
author: "Casey Connors"
date: "May 25, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Introduction

The goal of this project is to predict the manner in which test subjects performed 
a dumbbell exercise, the Unilateral Dumbbell Biceps Curl, using a set of data generated 
by accelerometers worn on the belt, forearm, arm, and dumbell of 6 participants. The 
data is tested on a set of 20 test cases from the same 6 participants.

Each participant performed 10 repetitions in five different ways.

- Class A: Exactly according to the specification of the exercise.
- Class B: Throwing the elbows to the front.
- Class C: Lifting the dumbbell only halfway.
- Class D: Lowering the dumbbell only halfway.
- Class E: Throwing the hips to the front.


#Data Exploration

```{r echo=TRUE}
library(caret)

training <- read.csv("data/pml-training.csv")
testing <- read.csv("data/pml-testing.csv")

str(training)

str(training$classe)



```


#Regression Prediction Model

```{r, echo=TRUE}
cvCtrlTime <- trainControl(method = "timeslice", initialWindow = 10, horizon = 1, fixedWindow = FALSE) #Will conduct time slice cross-validation on the next sample, growing the sample from an initial size of 10, always starting and 1.

cvCtrlKfold <- trainControl(method = "repeatedcv", repeats = 3) #will conduct a 10-fold repeated cross-validation three times.


#train the model


#test the model - predict then look at confusion matrix.


```

#KNN Model


#SVM model


#Random Forest Prediction Model


#Results and Conclusion